### 2024-02-20 paper_sisc_first_revision/localpsf.tex
L -> r_0 in RBF to resolve notation conflict

If there are negative numbers in the integral kernel (violating Assumption 3), the ellipsoid estimation procedure may incur errors or fail, leading to poor performance or failure of the PSF-based method. In Figure \ref{fig:robustness_to_negatives} we investigate the robustness of the ellipsoid support estimation procedure to violations of Assumption 3. We study two integral kernel examples, both of which are parameterized by a quantity that controls how negative the kernels are. We make the following observations:
\begin{itemize}
	\item The larger and more numerous the negative numbers are, the more inaccurate the ellipsoid support estimate is.
	\item The further away from the center of the ellipsoid the negative numbers are, the more influence they have on the quality of the ellipsoid support estimate. This is because moment formulas (Equations \ref{eq:define_mean} and \ref{eq:define_cov}) assign more weight to entries in the kernel that are further from the center.
	\item Negative numbers in the integral kernel affect the ellipsoid estimation method more if they are isolated, and less if they are balanced by nearby positive numbers.
	\item As kernels become more negative, the ellipsoid estimation performs well up to a certain threshold that depends on the spatial distribution of negative and positive entries. After that threshold is crossed, the estimation rapidly transitions to performing poorly and ultimately failing.
\end{itemize} 
%The location of the negative numbers also affects the accuracy of the ellipsoid estimates substantially, but in a more complicated manner. 
For the kernel in the left two columns of Figure \ref{fig:robustness_to_negatives}, negative numbers are interspersed with positive numbers, allowing us to include a large amount of negative numbers before the ellipsoid estimation fails. For the kernel in the right two columns, the ellipsoid estimate fails with tiny amounts of negative numbers because the negative numbers are far away from the mean and not balanced by positive numbers at similar distances and angles. In the bottom two rows, we see the aforementioned threshold effect, in which the ellipsoid estimation method rapidly transitions from performing reasonably well to performing poorly with only a small change to the integral kernel.

@@

The impulse response $\impulseresponse_x$ may be interpreted as a scaled probability distribution because of the non-negative integral kernel property. Let $\spatialvol:\Omega \rightarrow \mathbb{R}$,
%
%
\begin{equation}
\label{eq:define_vol}
\spatialvol(x) := \int_\Omega \impulseresponse_x(y) \mathrm{d}y,
\end{equation}
%
%
denote the spatially varying scaling factor, and for $i,j=1,\dots,\gdim$ define $\spatialmean:\Omega \rightarrow \mathbb{R}^\gdim$ and $\spatialcov:\Omega \rightarrow \mathbb{R}^{\gdim \times \gdim}$ as follows:
%
%
\begin{align}
\spatialmean^i(x) :=& \frac{1}{V(x)}\int_\Omega \impulseresponse_x(y) y^i ~\mathrm{d}y \label{eq:define_mean} \\
\spatialcov^{ij}(x) :=& \frac{1}{V(x)}\int_\Omega \impulseresponse_x(y) \left(y^i - \spatialmean^i(x)\right) \left(y^j - \spatialmean^j(x)\right) ~\mathrm{d}y, \label{eq:define_cov}
\end{align}
%
%
where $\spatialmean^i(x)$ and $y^i$ denote the $i^\text{th}$ components of the vectors $\spatialmean(x)$ and $y$, respectively, and $\spatialcov^{ij}(x)$ denotes the $(i,j)$ entry of the matrix $\spatialcov(x)$.
The vector $\spatialmean(x)\in \mathbb{R}^\gdim$ and the matrix $\spatialcov(x) \in \mathbb{R}^{\gdim \times \gdim}$ are the mean and covariance of the normalized version of $\impulseresponse_x$, respectively.

@@

Given $(y,x)\in \Omega \times \Omega$, let
$
	z_i := y - \spatialmean(x) + \spatialmean(x_i)
$
and define
\begin{equation}
\label{eq:fxyxp}
f_i := \frac{\spatialvol(x)}{\spatialvol(x_i)}\phi_{x_i}\left(z_i\right)
\end{equation}
for $i=1,\dots,\numnbr$, where $\{x_i\}_{i=1}^{\numnbr}$ are the $\numnbr$ nearest sample points to $x$, excluding points $x_i$ for which $z_i \notin \Omega$. Here, $\numnbr$ is a small user-defined parameter, e.g., $\numnbr=10$. We find the $\numnbr$ nearest sample points to $x$ by querying a precomputed kd-tree~\cite{Bentley75} of all sample points.  We check whether $z_i \in \Omega$ by querying a precomputed axis aligned bounding box tree (AABB tree)~\cite{Ericson04} of the mesh cells used to discretize the problem. 
Note that $\phi_{x_i}\left(z_i\right)$ is well-defined because $z_i \in \Omega$, and $\frac{\spatialvol(x)}{\spatialvol(x_i)}$ is well-defined because the sample point choosing procedure in Section~\ref{sec:sample_point_selection} ensures that $\spatialvol(x_i)>0$. Per the discussion in Section~\ref{sec:local_mean_displacement_invariance}, we expect $\Aker(y,x) \approx f_i$ for $i=1,\dots,\numnbr$. The closer $x_i$ is to $x$, the better we expect the approximation to be. We therefore approximate $\Aker(y,x)$ by interpolating the (point,value) pairs
$
\left\{\left(x_i, f_i\right)\right\}_{i=1}^{\numnbr}
$
at the point $x$. 
Interpolation is performed using the following radial basis function~\cite{Wendland04} scheme:
\begin{equation}
\Aker(y,x) \approx \widetilde{\Aker}(y,x) := \sum_{i=1}^{\numnbr} \rbfweight_i~ \rbf\left(\|x-x_i\|\right),
\end{equation}
where $\rbfweight_i$ are weights, and 
$\rbf(r):= \exp\left(-\frac{1}{2}\left(\epsilon\frac{r}{L}\right)^2\right)$
is a Gaussian kernel radial basis function. Here, $L:=\diam\left(\{x_i\}_{i=1}^{\numnbr}\right)$ is the diameter of the set of sample points used in the interpolation, and $\epsilon$ is a user-defined shape parameter that controls the width of the kernel function. The vector of weights, $\rbfweight = (\rbfweight_1, \rbfweight_2, \dots, \rbfweight_{\numnbr})^T$, is found as the solution to the $\numnbr \times \numnbr$ linear system
\begin{equation}
B \rbfweight = f,
\end{equation}
where $B \in \mathbb{R}^{\numnbr \times \numnbr}$,
$
B_{ij} := \rbf\left(\|x_i - x_j\|\right),
$
and $f \in \mathbb{R}^{\numnbr}$ has entries $f_i$ from~\eqref{eq:fxyxp}.

@@

The shape parameter, $\epsilon$, mediates a tradeoff between accuracy and stability. Small $\epsilon$ is required for RBF interpolation with Gaussian kernels to achieve high accuracy, but small $\epsilon$ also makes RBF interpolation less robust to errors or nonsmoothness in the function being interpolated. 
%This is illustrated in Figure \ref{fig:RBF_width} for a simple one dimensional interpolation problem. 
For our numerical results involving Hessians in inverse problems governed by PDEs (Sections~\ref{sec:ice}~and~\ref{sec:adv}), high accuracy is not required because the PSF-based method is used to construct a preconditioner. Hence, for these Hessian approximations we use a conservative choice of $\epsilon=3.0$ to ensure robustness. For our numerical results for the blur problem example (Section~\ref{sec:frog}), we use a smaller value of $\epsilon=0.5$ so that the RBF interpolation accuracy is not a limiting factor as we study convergence of the PSF-based method.

@@

In Table~\ref{tab:frog_psf_vs_hodlr_vs_glr} we compare the
computational cost to approximate the blur kernel from
Equation~\eqref{eq:frog_kernel} using the PSF-based method, the
randomized HODLR (hierarchical off diagonal low rank)
method~\cite{Martinsson16,Hartland_2023} with 8 hierarchical levels, and GLR (global low rank)
approximation using double-pass randomized SVD~\cite{HalkoMartinssonTropp11}. For
these results we vary the quantity $L$ to scale the width of the
impulse responses and hence the rank of the operator. For each case,
we calculate the relative error in the approximation of the kernel
measured in the Frobenius norm, $||\boldsymbol{\Phi} -
\widetilde{\boldsymbol{\Phi}}||_\text{Fro} /
||\boldsymbol{\Phi}||_\text{Fro}$ and show the number of operator
applications required to achieve 20\%, 10\%, and 5\% relative error by
each method. The results reveal superior performance of the PSF
method as compared to the HODLR and GLR methods for all cases. We note
that as we increase the rank and decrease the error tolerance, the
performance of the HODLR and GLR methods deteriorates.

>>>>

If there are negative numbers in the integral kernel (violating Assumption 3), the ellipsoid estimation procedure may incur errors or fail, leading to poor performance or failure of the PSF-based method. In Figure \ref{fig:robustness_to_negatives} we investigate the robustness of the ellipsoid support estimation procedure to violations of Assumption 3. We study two integral kernel examples, both of which are parameterized by a quantity that controls how negative the kernels are. We make the following observations:
\begin{itemize}
	\item The larger and more numerous the negative numbers are, the more inaccurate the ellipsoid support estimate is.
	\item The further away from the center of the ellipsoid the negative numbers are, the more influence they have on the quality of the ellipsoid support estimate. This is because moment formulas (Equations \ref{eq:define_mean} and \ref{eq:define_cov}) assign more weight to entries in the kernel that are further from the center.
	\item Negative numbers affect the ellipsoid estimation method more if they are isolated, and less if they are balanced by nearby positive numbers.
	\item As kernels become more negative, the ellipsoid estimation performs well up to a certain threshold that depends on the spatial distribution of negative and positive entries. After that threshold is crossed, the estimation rapidly transitions to performing poorly and ultimately failing.
\end{itemize} 
%The location of the negative numbers also affects the accuracy of the ellipsoid estimates substantially, but in a more complicated manner. 
For the kernel in the left two columns of Figure \ref{fig:robustness_to_negatives}, negative numbers are interspersed with positive numbers, allowing us to include a large amount of negative numbers before the ellipsoid estimation fails. For the kernel in the right two columns, the ellipsoid estimate fails with tiny amounts of negative numbers because the negative numbers are far away from the mean and not balanced by positive numbers at similar distances and angles. In the bottom two rows, we see the aforementioned threshold effect, in which the ellipsoid estimation method rapidly transitions from performing reasonably well to performing poorly with only a small change to the integral kernel.

@@

The impulse response $\impulseresponse_x$ may be interpreted as a scaled probability distribution because of the non-negative integral kernel property. Let $\spatialvol:\Omega \rightarrow \mathbb{R}$,
%
%
\begin{equation}
\label{eq:define_vol}
\spatialvol(x) := \int_\Omega \impulseresponse_x(y) \mathrm{d}y,
\end{equation}
%
%
denote the spatially varying scaling factor, and for $i,j=1,\dots,\gdim$ define $\spatialmean:\Omega \rightarrow \mathbb{R}^\gdim$ and $\spatialcov:\Omega \rightarrow \mathbb{R}^{\gdim \times \gdim}$ as follows:
%
%
\begin{align}
\spatialmean^i(x) :=& \frac{1}{V(x)}\int_\Omega \impulseresponse_x(y) y^i ~\mathrm{d}y \label{eq:define_mean} \\
\spatialcov^{ij}(x) :=& \frac{1}{V(x)}\int_\Omega \impulseresponse_x(y) \left(y^i - \spatialmean^i(x)\right) \left(y^j - \spatialmean^j(x)\right) ~\mathrm{d}y, \label{eq:define_cov}
\end{align}
%
%
where $\spatialmean^i(x)$ and $y^i$ denote the $i^\text{th}$ components of the vectors $\spatialmean(x)$ and $y$, respectively, and $\spatialcov^{ij}(x)$ denotes the $(i,j)$ entry of the matrix $\spatialcov(x)$.
The quantities $\spatialmean(x)\in \mathbb{R}^\gdim$ and $\spatialcov(x) \in \mathbb{R}^{\gdim \times \gdim}$ are the mean and covariance of the normalized version of $\impulseresponse_x$, respectively.

@@

Given $(y,x)\in \Omega \times \Omega$, let
$
	z_i := y - \spatialmean(x) + \spatialmean(x_i)
$
and define
\begin{equation}
\label{eq:fxyxp}
f_i := \frac{\spatialvol(x)}{\spatialvol(x_i)}\phi_{x_i}\left(z_i\right)
\end{equation}
for $i=1,\dots,\numnbr$, where $\{x_i\}_{i=1}^{\numnbr}$ are the $\numnbr$ nearest sample points to $x$, excluding points $x_i$ for which $z_i \notin \Omega$. Here, $\numnbr$ is a small user-defined parameter, e.g., $\numnbr=10$. We find the $\numnbr$ nearest sample points to $x$ by querying a precomputed kd-tree~\cite{Bentley75} of all sample points.  We check whether $z_i \in \Omega$ by querying a precomputed axis aligned bounding box tree (AABB tree)~\cite{Ericson04} of the mesh cells used to discretize the problem. 
Note that $\phi_{x_i}\left(z_i\right)$ is well-defined because $z_i \in \Omega$, and $\frac{\spatialvol(x)}{\spatialvol(x_i)}$ is well-defined because the sample point choosing procedure in Section~\ref{sec:sample_point_selection} ensures that $\spatialvol(x_i)>0$. Per the discussion in Section~\ref{sec:local_mean_displacement_invariance}, we expect $\Aker(y,x) \approx f_i$ for $i=1,\dots,\numnbr$. The closer $x_i$ is to $x$, the better we expect the approximation to be. We therefore approximate $\Aker(y,x)$ by interpolating the (point,value) pairs
$
\left\{\left(x_i, f_i\right)\right\}_{i=1}^{\numnbr}
$
at the point $x$. 
Interpolation is performed using the following radial basis function~\cite{Wendland04} scheme:
\begin{equation}
\Aker(y,x) \approx \widetilde{\Aker}(y,x) := \sum_{i=1}^{\numnbr} \rbfweight_i~ \rbf\left(\|x-x_i\|\right),
\end{equation}
where $\rbfweight_i$ are weights, and 
$\rbf(r):= \exp\left(-\frac{1}{2}\left(C_\text{RBF}\frac{r}{r_0}\right)^2\right)$
is a Gaussian kernel radial basis function. Here, $r_0:=\diam\left(\{x_i\}_{i=1}^{\numnbr}\right)$ is the diameter of the set of sample points used in the interpolation, and $C_\text{RBF}$ is a user-defined shape parameter that controls the width of the kernel function. The vector of weights, $\rbfweight = (\rbfweight_1, \rbfweight_2, \dots, \rbfweight_{\numnbr})^T$, is found as the solution to the $\numnbr \times \numnbr$ linear system
\begin{equation}
B \rbfweight = f,
\end{equation}
where $B \in \mathbb{R}^{\numnbr \times \numnbr}$,
$
B_{ij} := \rbf\left(\|x_i - x_j\|\right),
$
and $f \in \mathbb{R}^{\numnbr}$ has entries $f_i$ from~\eqref{eq:fxyxp}.

@@

The shape parameter, $C_\text{RBF}$, mediates a tradeoff between accuracy and stability. Small $C_\text{RBF}$ is required for RBF interpolation with Gaussian kernels to achieve high accuracy, but small $C_\text{RBF}$ also makes RBF interpolation less robust to errors or nonsmoothness in the function being interpolated. 
%This is illustrated in Figure \ref{fig:RBF_width} for a simple one dimensional interpolation problem. 
For our numerical results involving Hessians in inverse problems governed by PDEs (Sections~\ref{sec:ice}~and~\ref{sec:adv}), high accuracy is not required because the PSF-based method is used to construct a preconditioner. Hence, for these Hessian approximations we use a conservative choice of $C_\text{RBF}=3.0$ to ensure robustness. For our numerical results for the blur problem example (Section~\ref{sec:frog}), we use a smaller value of $C_\text{RBF}=0.5$ so that the RBF interpolation accuracy is not a limiting factor as we study convergence of the PSF-based method.

@@

In Table~\ref{tab:frog_psf_vs_hodlr_vs_glr} we compare the
cost to approximate the blur kernel from
Equation~\eqref{eq:frog_kernel} using the PSF-based method, the
randomized HODLR (hierarchical off diagonal low rank)
method~\cite{Martinsson16,Hartland_2023} with 8 levels, and GLR (global low rank)
approximation using double-pass randomized SVD~\cite{HalkoMartinssonTropp11}. For
these results we vary the quantity $L$ to scale the width of the
impulse responses and hence the rank of the operator. For each case,
we calculate the relative error in the approximation of the kernel
measured in the Frobenius norm, $||\boldsymbol{\Phi} -
\widetilde{\boldsymbol{\Phi}}||_\text{Fro} /
||\boldsymbol{\Phi}||_\text{Fro}$ and show the number of operator
applications required to achieve 20\%, 10\%, and 5\% relative error by
each method. The results reveal superior performance of the PSF
method as compared to the HODLR and GLR methods for all cases. We note
that as we increase the rank and decrease the error tolerance, the
performance of the HODLR and GLR methods deteriorates.
